The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-10-11 pkg/yuimadocs/inst/doc/JSS/final-edited.Rnw
edited for JSS final
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@251 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\bea{\begin{eqnarray}}
\def\eea{\end{eqnarray}}
\def\beas{\begin{eqnarray*}}
\def\eeas{\end{eqnarray*}}
\def\sskip{\hspace{5mm}}

@@

\section{The \pkg{yuima} Package}\label{sec2}
The package \pkg{yuima} depends on some other packages, like  \pkg{zoo} \citep{zoo}, which can be installed separately.
The package \pkg{zoo} is used internally to store time series data. This dependence may change in the future adopting a more flexible class for internal storage of time series.

\subsection{How to Obtain the Package}
The \pkg{yuima} package is hosted on R-Forge and the web page
of the Project is  \url{http://r-forge.r-project.org/projects/yuima}.
The  R-Forge page contains the latest development version, and stable
version of the package will also be available through CRAN.
Development versions of the package are not supposed to be stable or functional, thus
the occasional user should consider to install the stable version first.
The package can be installed from R-Forge using
\code{install.packages("yuima",repos="http://R-Forge.R-project.org")}.
If, the R-Forge system does not provide binary builds of the \pkg{yuima} package, the
user can also try 
\code{install.packages("yuima",repos="http://R-Forge.R-project.org", type="source")}.

\subsection{The Main Object and Classes}
Before discussing the methods for simulation and inference for stochastic processes solutions to stochastic differential equations, here we discuss the main classes in the package.
As mentioned there are different classes of objects defined in the \pkg{yuima} package and
the main class is called the \code{yuima-class}. This class  is composed of several slots.
Figure \ref{fig:classes} represents the different classes and their slots.
\begin{figure}[th]
\centering{\includegraphics[width=\textwidth]{yuima-class}}
\caption{The main classes in the \pkg{yuima} package.}
\label{fig:classes}
\end{figure} 
The different slots do not need to be all present at the same time. For example, in case one wants to simulate a stochastic process, only the slots \code{model} and \code{sampling} should be present, while the slot \code{data} will be filled by the simulator.
We discuss in detail the different objects separately in the next sections.

@@

\subsection{The \code{yuima.model} Class}\label{sec:model}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified.  Here we present  a brief
overview of these models as they will be described in detail in Section \ref{sec3}, but this allows to introduce an overall view of the slots of the \code{yuima.model} class.
In \pkg{yuima} one can describe three main families of stochastic processes at present. These models can be one or multidimensional and eventually described as parametric models. Let $X_0=x_0$ be the initial value of the process, then, the main classes are as follows:
\begin{itemize}
\item diffusion models described as
$$  \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t, \quad X_0=x_0$$
 where $W_t$ is a standard Brownian motion;
\item SDE's driven by fractional Gaussian noise, with $H$ the Hurst parameter, described as
$$ \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t^{H}, \quad X_0=x_0;$$ 
\item diffusion process with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0;
$$
\end{itemize}
 The functions $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ may have a different number of arguments. For example, if the model is homogeneous in time and drift and diffusion coefficients are entirely specified, then we will use the notaion $a(x)$ and $b(x)$ and describe the diffusion model simply as $\displaystyle \de X_t=a(X_t)dt + b(X_t)\de W_t$. And so forth. Detailed hypotheses and regularity conditions on the coefficients $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ for each class of models will be given in the next sections. Nevertheless, it is important to remark that these notations only matter to the mathematical description of the model while each coefficient is passed to \pkg{yuima} methods as \proglang{R} mathematical expressions. This means that, for example, $a(t, X_t, \theta) = t\cdot \sqrt{\theta X_t}$ will be passed as \code{t*sqrt(x*theta)}, therefore, the order of the arguments is not relevant to \proglang{R} as well as its mathematical description as long as it is consistent through each specific section. Further, \pkg{yuima} is able to accept any user-specified notation for the state variable $x$ (for $X_t$) and the time variable $t$ and the remaining term of an \proglang{R} expression will be interpreted as parameter as will be explained in Section \ref{sec:diff}.
We are now able to give an overview of the main slots of the most important class of the \pkg{yuima} package.

@@

\section{Model Specification}\label{sec3}
In order to show how general the approach of the \pkg{yuima} package is, we present some examples. Throughout this section we assume that all the stochastic differential equations exist while in Section \ref{sec5} we will give  regularity conditions needed to have a properly defined statistical model.

\subsection{One Dimensional Diffusion Processes}\label{sec:diff}
Assume that we want to describe the following stochastic differential equation
$$\de X_t = -3 X_t \de t + \frac{1}{1+X_t^2}\de W_t.$$
In the above $a(x) = -3 x$ and $b(x) = \frac{1}{1+x^2}$ according to the notation of previous section and $W_t$ is a standard Wiener process.
This can be described in \pkg{yuima} by specifying the drift and diffusion coefficients as plain
\proglang{R} expressions passed as strings
<<print=FALSE,echo=FALSE,results=hide>>=
library(yuima)
@ 
<<echo=TRUE, print=FALSE,results=hide>>=
mod1 <- setModel(drift = "-3*x", diffusion = "1/(1+x^2)")
@ 
By default, \pkg{yuima} assumes that the state variable is \code{x} and the time variable is \code{t} and the solve variable is again \code{x}. Notice that the left hand-side of the equation is implicit, this is why \code{yuima.model} has the slot \code{solve.variable}.
The user should not be worried about the warning raised by \pkg{yuima} at this stage, as this is just to inform her on the implicit assumption on the solution variable.
At this point, the package fills the proper slots of the \code{yuima} object
<<>>=
str(mod1)
@

From the above, it is possible to see that the jump coefficient is void and the Hurst parameter is set to 0.5, because this is a model where the driving process is the standard Brownian motion, i.e. a fractional Brownian motion if Hurst index $H=\frac12$.
Now, with \code{mod1} in hands, it is extremely easy to simulate a trajectory by  Euler-Maruyama scheme of the process as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod1)
plot(X)
@

@@

\subsection{User Specified State and Time Variables}
Suppose now that the user wants to specify her own model using a prescribed notation, e.g., some SDE's like
$$\de Y_s = -3 s Y_s \de s + \frac{1}{1+Y_s^2}\de W_s,$$
where $a(s,y) = -3sy$ and $b(y) = 1/(1+y^2)$.
Then this model can be described in \pkg{yuima} as follows
<<echo=TRUE, print=FALSE,results=hide>>=
mod1b <- setModel(drift = "-3*s*y", diffusion = "1/(1+y^2)", 
state.var="y", time.var="s")
@
In this case the solution variable is the same as the state variable. Indeed, the \code{yuima.model} object appears as follows
<<>>=
str(mod1b)
@
Once again, the user may use the \code{simulate} method to perform simulation.

\subsection{Specification of Parametric Models}\label{sec:par}
Assume now that we want to specify a parametric model like this
$$\de X_t = -\theta X_t \de t + \frac{1}{1+X_t^\gamma}\de W_t$$
where $a(x,\theta) = -\theta x$ and $b(x,\gamma) = 1/(1+x^\gamma)$.
Then, \pkg{yuima} attempts to distinguish the parameters' names from the ones of the state and time variables
<<echo=TRUE, print=FALSE,results=hide>>=
mod2 <- setModel(drift = "-theta*x", diffusion = "1/(1+x^gamma)")
@
so, in this case, \code{theta} and \code{gamma}, which are different form \code{x} and \code{t}, are assumed to be parameters. Notice that in the above notation $\theta$ and $\gamma$ are generic names for the components of a parameters' vector $\theta$ in the notation of Section \ref{sec:model}.
<<>>=
str(mod2)
@
In order to simulate the parametric model it is necessary to specify the values of the parameters $\theta$ and $\gamma$ as shown in the next code chunk
<<echo=TRUE, print=FALSE,fig=TRUE,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod2,true.param=list(theta=1,gamma=3))
plot(X)
@

\subsection{Multidimensional Processes}\label{sec:multi}
Next is an example of a system of two stochastic differential equations for the couple $(X_{1,t}, X_{2,t})$ driven by three independent Brownian motions $(W_{1,t}, W_{2,t}, W_{3,t})$
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form with a vector of drift expressions and a diffusion matrix
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left(\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right)
\left(\begin{array}{c}\de W_{1,t} \\ \de W_{2,t} \\ \de W_{3,t}\end{array}\right)
$$
For this system it is now necessary to instruct \pkg{yuima} about the state variable on both the left-hand side of the equation and the right-hand side of the equation, i.e. there is the need to specify also the \code{solve.variable} for the left hand side of the SDE:
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.
Again, this model can be easily simulated
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@

Notice that the role of \code{solve.variable} is essential because it allows to specify the left hand side of an SDE equation. For example, if one wants to specify this model instead of the previous one

$$
\begin{aligned}
\de X_{2,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{1,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
the \code{solve.variable} argument should be specified as \code{solve.variable=c("x2","x1")} in place of  \code{solve.variable=c("x1","x2")}, all the rest being the same as in model \code{mod3}.

@@

\subsection{Fractional Gaussian Noise}\label{sec:fgn}
The \pkg{yuima} allows for the description of stochastic differential equations driven by fractional Brownian motion of the following type
$$\de X_t =  a(X_t) \de t + b(X_t) \de W_t^H
$$
where $W^H=\left( W^H_t, t\geq 0 \right)$ is a normalized fractional Brownian motion (fBM), {\it i.e.} the zero mean Gaussian processes with covariance function
$$\dE( W_s^H W_t^H )= \frac{1}{2} \left( |s|^{2H}+|t|^{2H}- |t-s|^{2H}\right)$$ with Hurst exponent $H\in(0,1)$.  The fractional Brownian motion process is  neither Markovian nor a semimartingale for $H\neq\frac{1}{2}$ but
remains Gaussian.  For $H > \frac12$, the solution $X_t$ above  presents the long-range dependence property that makes it useful for different applications in biology, physics, ethernet traffic or finance.
In order to specify a stochastic differential equation driven by fractional Gaussian noise it is necessary to specify the value of the Hurst parameter. For example, if we want to specify the following model
$$\de Y_t =  3 Y_t \de t + \de W_t^H$$
we proceed as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=6,results=hide>>=
mod4A <- setModel(drift="3*y", diffusion=1, hurst=0.3, solve.var="y")
mod4B <- setModel(drift="3*y", diffusion=1, hurst=0.7, solve.var="y")
set.seed(123)
X1 <- simulate(mod4A,sampling=setSampling(n=1000))
X2 <- simulate(mod4B,sampling=setSampling(n=1000))
par(mfrow=c(2,1))
par(mar=c(2,3,1,1))
plot(X1,main="H=0.3")
plot(X2,main="H=0.7")
@

@@

\subsection{L\'evy Processes}\label{sec:jump}
Jump processes can be specified in different ways in mathematics and hence in \pkg{yuima} package. 
Let $Z_t$ be a  Compound Poisson Process (i.e., jumps size follow some distribution, like the Gaussian law and jumps occur at Poisson times).
Then it is possible to consider the simple following SDE which involves jumps
$$\de X_t =  a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \de Z_t,$$
or the more general representation
\begin{equation}\label{eq:levy}
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0,
\end{equation}
with $\mu$ a random measure associated with the jumps of $X$, that is,
$$
\mu(dt,dz)=\sum_{s>0}\mathbf{1}_{\{\Delta Z_s \ne 0\}}\delta_{(s,\Delta Z_s)}(dt,dz),
$$
where $\delta$ denotes the Dirac measure and $Z$ the driving pure-jump L\'evy process of the form
$$
Z_{t}=\int_{0}^{t}\int_{|z|\le 1}z\{\mu(ds,dz) - \nu(dz)ds\}+\int_{0}^{t}\int_{|z|>1}z\mu(ds,dz).
$$
The L\'evy measure $\nu$ is any measure satisfying $\nu(\{0\})=0$ and $\int(1\wedge |z|^2)\nu(dz)< \infty$.
In the next example we consider a compound Poisson process with intensity $\lambda=10$ with Gaussian jumps as driving L\'evy process.
This model can be specified in \code{setModel} using the argument  \code{measure.type="CP"} (fro Compound Poisson).
A simple Ornstein-Uhlenbeck process with Gaussian jumps like this
$$\de X_t = -\theta X_t \de t + \sigma \de W_t + \de Z_t$$
is then specified as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod5 <- setModel(drift=c("-theta*x"), diffusion="sigma",
 jump.coeff="1", measure=list(intensity="10", df=list("dnorm(z, 0, 1)")),
 measure.type="CP", solve.variable="x")
set.seed(123)
X <- simulate(mod5, true.p=list(theta=1,sigma=3),sampling=setSampling(n=1000))
plot(X)
@

\noindent
Another possibility is to specify the jump component via the L\'evy measure $\nu(\cdot)$. 
 Without going into too much details, here is an example of specification of a simple Ornstein-Uhlenbeck  process with IG (Inverse Gaussian) L\'evy measure $\nu(\cdot)$ and no Poisson component
$$\de X_t = -x \de t + \de Z_t$$
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod6 <- setModel(drift="-x", xinit=1, jump.coeff="1", 
  measure.type="code", measure=list(df="rIG(z, 1, 0.1)"))
set.seed(123)
X <- simulate(mod6, sampling=setSampling(Terminal=10, n=10000)) 
plot(X)
@

>>>>

\usepackage{float}
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\bea{\begin{eqnarray}}
\def\eea{\end{eqnarray}}
\def\beas{\begin{eqnarray*}}
\def\eeas{\end{eqnarray*}}
\def\sskip{\hspace{5mm}}

@@

\section{The \pkg{yuima} package}\label{sec2}
The package \pkg{yuima} depends on some other packages, like  \pkg{zoo} \citep{zoo}, which can be installed separately.
The package \pkg{zoo} is used internally to store time series data. This dependence may change in the future adopting a more flexible class for internal storage of time series.

\subsection{How to obtain the package}
The \pkg{yuima} package is hosted on R-Forge and the web page
of the Project is  \url{http://r-forge.r-project.org/projects/yuima}.
The  R-Forge page contains the latest development version, and stable
version of the package will also be available through CRAN.
Development versions of the package are not supposed to be stable or functional, thus
the occasional user should consider to install the stable version first.
The package can be installed from R-Forge using
\code{install.packages("yuima",repos="http://R-Forge.R-project.org")}.
If, the R-Forge system does not provide binary builds of the \pkg{yuima} package, the
user can also try 
\code{install.packages("yuima",repos="http://R-Forge.R-project.org", type="source")}.

\subsection{The main object and classes}
Before discussing the methods for simulation and inference for stochastic processes solutions to stochastic differential equations, here we discuss the main classes in the package.
As mentioned there are different classes of objects defined in the \pkg{yuima} package and
the main class is called the \code{yuima-class}. This class  is composed of several slots.
Figure~\ref{fig:classes} represents the different classes and their slots.
\begin{figure}[th]
\centering{\includegraphics[width=\textwidth]{yuima-class}}
\caption{The main classes in the \pkg{yuima} package.}
\label{fig:classes}
\end{figure} 
The different slots do not need to be all present at the same time. For example, in case one wants to simulate a stochastic process, only the slots \code{model} and \code{sampling} should be present, while the slot \code{data} will be filled by the simulator.
We discuss in detail the different objects separately in the next sections.

@@

\subsection{The \code{yuima.model} class}\label{sec:model}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified.  Here we present  a brief
overview of these models as they will be described in detail in Section \ref{sec3}, but this allows to introduce an overall view of the slots of the \code{yuima.model} class.
In \pkg{yuima} one can describe three main families of stochastic processes at present. These models can be one or multidimensional and eventually described as parametric models. Let $X_0=x_0$ be the initial value of the process, then, the main classes are as follows:
\begin{itemize}
\item diffusion models described as
$$  \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t, \quad X_0=x_0$$
 where $W_t$ is a standard Brownian motion;
\item SDE's driven by fractional Gaussian noise, with $H$ the Hurst parameter, described as
$$ \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t^{H}, \quad X_0=x_0;$$ 
\item diffusion process with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0;
$$
\end{itemize}
 The functions $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ may have a different number of arguments. For example, if the model is homogeneous in time and drift and diffusion coefficients are entirely specified, then we will use the notaion $a(x)$ and $b(x)$ and describe the diffusion model simply as $\displaystyle \de X_t=a(X_t)dt + b(X_t)\de W_t$. And so forth. Detailed hypotheses and regularity conditions on the coefficients $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ for each class of models will be given in the next sections. Nevertheless, it is important to remark that these notations only matter to the mathematical description of the model while each coefficient is passed to \pkg{yuima} methods as \proglang{R} mathematical expressions. This means that, for example, $a(t, X_t, \theta) = t\cdot \sqrt{\theta X_t}$ will be passed as \code{t*sqrt(x*theta)}, therefore, the order of the arguments is not relevant to \proglang{R} as well as its mathematical description as long as it is consistent through each specific section. Further, \pkg{yuima} is able to accept any user-specified notation for the state variable $x$ (for $X_t$) and the time variable $t$ and the remaining term of an \proglang{R} expression will be interpreted as parameter as will be explained in Section \ref{sec:diff}.
We are now able to give an overview of the main slots of the most important class of the \pkg{yuima} package.

@@

\section{Model specification}\label{sec3}
In order to show how general the approach of the \pkg{yuima} package is, we present some examples. Throughout this section we assume that all the stochastic differential equations exist while in Section \ref{sec5} we will give  regularity conditions needed to have a properly defined statistical model.

\subsection{One dimensional diffusion processes}\label{sec:diff}
Assume that we want to describe the following stochastic differential equation
$$\de X_t = -3 X_t \de t + \frac{1}{1+X_t^2}\de W_t.$$
In the above $a(x) = -3 x$ and $b(x) = \frac{1}{1+x^2}$ according to the notation of previous section and $W_t$ is a standard Wiener process.
This can be described in \pkg{yuima} by specifying the drift and diffusion coefficients as plain
\proglang{R} expressions passed as strings
<<print=FALSE,echo=FALSE,results=hide>>=
library("yuima")
@ 
<<echo=TRUE, print=FALSE,results=hide>>=
mod1 <- setModel(drift = "-3*x", diffusion = "1/(1+x^2)")
@ 
By default, \pkg{yuima} assumes that the state variable is \code{x} and the time variable is \code{t} and the solve variable is again \code{x}. Notice that the left hand-side of the equation is implicit, this is why \code{yuima.model} has the slot \code{solve.variable}.
The user should not be worried about the warning raised by \pkg{yuima} at this stage, as this is just to inform her on the implicit assumption on the solution variable.
At this point, the package fills the proper slots of the \code{yuima} object
<<>>=
str(mod1)
@

From the above, it is possible to see that the jump coefficient is void and the Hurst parameter is set to 0.5, because this is a model where the driving process is the standard Brownian motion, i.e., a fractional Brownian motion if Hurst index $H=\frac12$.
Now, with \code{mod1} in hands, it is extremely easy to simulate a trajectory by  Euler-Maruyama scheme of the process as follows
<<sim-mod1,echo=TRUE, print=FALSE,results=hide>>=
set.seed(123)
X <- simulate(mod1)
@
which can be plotted using the command \code{plot}
<<plot-mod1,echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
plot(X)
@
and the result is shown in Figure~\ref{fig:plot-mod1}.
\begin{figure}[H]
<<echo=FALSE, print=FALSE,results=hide>>=
pdf("plot-mod1.pdf",width=9,height=4)
par(mar=c(4,4,0,0))
plot(X)
dev.off()
@
\includegraphics[width=\textwidth]{plot-mod1}
\caption{The \code{plot} function is used to draw a trajectory of a simulated \code{yuima} object.}
\label{fig:plot-mod1}
\end{figure}

@@

\subsection{User specified state and time variables}
Suppose now that the user wants to specify her own model using a prescribed notation, e.g., some SDE's like
$$\de Y_s = -3 s Y_s \de s + \frac{1}{1+Y_s^2}\de W_s,$$
where $a(s,y) = -3sy$ and $b(y) = 1/(1+y^2)$.
Then this model can be described in \pkg{yuima} as follows
<<echo=TRUE, print=FALSE,results=hide>>=
mod1b <- setModel(drift = "-3*s*y", diffusion = "1/(1+y^2)", 
state.var="y", time.var="s")
@
In this case the solution variable is the same as the state variable. Indeed, the \code{yuima.model} object appears as follows
<<>>=
str(mod1b)
@
Once again, the user may use the \code{simulate} method to perform simulation.

\subsection{Specification of parametric models}\label{sec:par}
Assume now that we want to specify a parametric model like this
$$\de X_t = -\theta X_t \de t + \frac{1}{1+X_t^\gamma}\de W_t$$
where $a(x,\theta) = -\theta x$ and $b(x,\gamma) = 1/(1+x^\gamma)$.
Then, \pkg{yuima} attempts to distinguish the parameters' names from the ones of the state and time variables
<<echo=TRUE, print=FALSE,results=hide>>=
mod2 <- setModel(drift = "-theta*x", diffusion = "1/(1+x^gamma)")
@
so, in this case, \code{theta} and \code{gamma}, which are different form \code{x} and \code{t}, are assumed to be parameters. Notice that in the above notation $\theta$ and $\gamma$ are generic names for the components of a parameters' vector $\theta$ in the notation of Section \ref{sec:model}.
<<>>=
str(mod2)
@
In order to simulate the parametric model it is necessary to specify the values of the parameters $\theta$ and $\gamma$ as shown in the next code chunk
<<sim-mod2,echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
set.seed(123)
X <- simulate(mod2,true.param=list(theta=1,gamma=3))
plot(X)
@
and the trajectory can be seen in Figure~\ref{fig:plot-mod2}.
\begin{figure}[H]
<<plot-mod2,echo=FALSE, print=FALSE,results=hide>>=
pdf("plot-mod2.pdf",width=9,height=4)
par(mar=c(4,4,0,0))
plot(X)
dev.off()
@
\includegraphics[width=\textwidth]{plot-mod2}
\caption{A trajectory simulated from the parametric model  \code{mod2}.}
\label{fig:plot-mod2}
\end{figure}

\subsection{Multidimensional processes}\label{sec:multi}
Next is an example of a system of two stochastic differential equations for the couple $(X_{1,t}, X_{2,t})$ driven by three independent Brownian motions $(W_{1,t}, W_{2,t}, W_{3,t})$
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form with a vector of drift expressions and a diffusion matrix
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left(\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right)
\left(\begin{array}{c}\de W_{1,t} \\ \de W_{2,t} \\ \de W_{3,t}\end{array}\right)
$$
For this system it is now necessary to instruct \pkg{yuima} about the state variable on both the left-hand side of the equation and the right-hand side of the equation, i.e., there is the need to specify also the \code{solve.variable} for the left hand side of the SDE:
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.
Again, this model can be easily simulated and the trajectory can be seen in Figure~\ref{fig:plot-mod3}.
<<sim-mod3,echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@
<<plot-mod3,echo=FALSE, fig=FALSE, print=FALSE,results=hide>>=
pdf("plot-mod3.pdf",width=9,height=4)
par(mar=c(4,4,0,0))
plot(X, plot.type="single",lty=1:2)
dev.off()
@
\begin{figure}[H]
\includegraphics[width=\textwidth]{plot-mod3}
\caption{A trajectory of the multidimensional SDE described in \code{mod3}.}
\label{fig:plot-mod3}
\end{figure} 
Notice that the role of \code{solve.variable} is essential because it allows to specify the left hand side of an SDE equation. For example, if one wants to specify this model instead of the previous one
$$
\begin{aligned}
\de X_{2,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{1,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
the \code{solve.variable} argument should be specified as \code{solve.variable=c("x2","x1")} in place of  \code{solve.variable=c("x1","x2")}, all the rest being the same as in model \code{mod3}.

@@

\subsection{Fractional gaussian noise}\label{sec:fgn}
The \pkg{yuima} allows for the description of stochastic differential equations driven by fractional Brownian motion of the following type
$$\de X_t =  a(X_t) \de t + b(X_t) \de W_t^H
$$
where $W^H=\left( W^H_t, t\geq 0 \right)$ is a normalized fractional Brownian motion (fBM), {\it i.e.}, the zero mean Gaussian processes with covariance function
$$\dE( W_s^H W_t^H )= \frac{1}{2} \left( |s|^{2H}+|t|^{2H}- |t-s|^{2H}\right)$$ with Hurst exponent $H\in(0,1)$.  The fractional Brownian motion process is  neither Markovian nor a semimartingale for $H\neq\frac{1}{2}$ but
remains Gaussian.  For $H > \frac12$, the solution $X_t$ above  presents the long-range dependence property that makes it useful for different applications in biology, physics, ethernet traffic or finance.
In order to specify a stochastic differential equation driven by fractional Gaussian noise it is necessary to specify the value of the Hurst parameter. For example, if we want to specify the following  fractional Ornstein-Uhlenbeck model
$$\de Y_t =  3 Y_t \de t + \de W_t^H$$
we can proceed as follows
<<sim-mod4AB, echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
mod4A <- setModel(drift="3*y", diffusion=1, hurst=0.3, solve.var="y")
mod4B <- setModel(drift="3*y", diffusion=1, hurst=0.7, solve.var="y")
set.seed(123)
X1 <- simulate(mod4A,sampling=setSampling(n=1000))
X2 <- simulate(mod4B,sampling=setSampling(n=1000))
par(mfrow=c(2,1))
par(mar=c(2,3,1,1))
plot(X1,main="H=0.3")
plot(X2,main="H=0.7")
@
and the two trajectories can be seen in Figure~\ref{fig:plot-mod4AB}.
\begin{figure}[H]
<<plot-mod4AB,echo=FALSE, print=FALSE,results=hide>>=
pdf("plot-mod4AB.pdf",width=9,height=6)
par(mfrow=c(2,1))
par(mar=c(2,3,1,1))
plot(X1,main="H=0.3")
plot(X2,main="H=0.7")
dev.off()
@
\includegraphics[width=\textwidth]{plot-mod4AB}
\caption{Trajectories of the fractional Ornstein-Uhlenbeck process for different values of the Hurst parameter.}
\label{fig:plot-mod4AB}
\end{figure}

@@

\subsection{L\'evy processes}\label{sec:jump}
Jump processes can be specified in different ways in mathematics and hence in \pkg{yuima} package. 
Let $Z_t$ be a  Compound Poisson Process (i.e., jumps size follow some distribution, like the Gaussian law and jumps occur at Poisson times).
Then it is possible to consider the simple following SDE which involves jumps
$$\de X_t =  a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \de Z_t,$$
or the more general representation
\begin{equation}\label{eq:levy}
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0,
\end{equation}
with $\mu$ a random measure associated with the jumps of $X$, that is,
$$
\mu(dt,dz)=\sum_{s>0}\mathbf{1}_{\{\Delta Z_s \ne 0\}}\delta_{(s,\Delta Z_s)}(dt,dz),
$$
where $\delta$ denotes the Dirac measure and $Z$ the driving pure-jump L\'evy process of the form
$$
Z_{t}=\int_{0}^{t}\int_{|z|\le 1}z\{\mu(ds,dz) - \nu(dz)ds\}+\int_{0}^{t}\int_{|z|>1}z\mu(ds,dz).
$$
The L\'evy measure $\nu$ is any measure satisfying $\nu(\{0\})=0$ and $\int(1\wedge |z|^2)\nu(dz)< \infty$.
In the next example we consider a compound Poisson process with intensity $\lambda=10$ with Gaussian jumps as driving L\'evy process.
This model can be specified in \code{setModel} using the argument  \code{measure.type="CP"} (fro Compound Poisson).
A simple Ornstein-Uhlenbeck process with Gaussian jumps like this
$$\de X_t = -\theta X_t \de t + \sigma \de W_t + \de Z_t$$
is then specified as follows (the trajectory is shown in Figure~\ref{fig:plot-mod5})
<<sim-mod5,echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
mod5 <- setModel(drift=c("-theta*x"), diffusion="sigma",
 jump.coeff="1", measure=list(intensity="10", df=list("dnorm(z, 0, 1)")),
 measure.type="CP", solve.variable="x")
X <- simulate(mod5, true.p=list(theta=1,sigma=3),sampling=setSampling(n=1000))
plot(X)
@
\begin{figure}[H]
<<plot-mod5,echo=FALSE, print=FALSE,results=hide>>=
pdf("plot-mod5.pdf",width=9,height=4)
par(mar=c(4,4,0,0))
plot(X)
dev.off()
@
\includegraphics[width=\textwidth]{plot-mod5}
\caption{A trajectory of the Ornstein-Uhlenbeck process with jumps following a compound Poisson process with Gaussian jumps as defined in \code{mod5}.}
\label{fig:plot-mod5}
\end{figure} 
%%
\noindent
Another possibility is to specify the jump component via the L\'evy measure $\nu(\cdot)$. 
 Without going into too much details, here is an example of specification of a simple Ornstein-Uhlenbeck  process with IG (Inverse Gaussian) L\'evy measure $\nu(\cdot)$ and no Poisson component (see Figure~\ref{fig:plot-mod6})
$$\de X_t = -x \de t + \de Z_t$$
<<sim-mod6,echo=TRUE, print=FALSE,fig=FALSE,results=hide>>=
mod6 <- setModel(drift="-x", xinit=1, jump.coeff="1", 
  measure.type="code", measure=list(df="rIG(z, 1, 0.1)"))
X <- simulate(mod6, sampling=setSampling(Terminal=10, n=10000)) 
plot(X)
@
\noindent
